Comparison of Several Stochastic and Deterministic Derivative-free Global Optimization Algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}

### 2019-02-14 2019/MOTOR_Sovrasov/motor_paper.tex
MOToR: pre-submit version

\begin{document}
%
\title{Comparison of several stochastic and deterministic derivative-free global optimization algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Vladislav Sovrasov}
%
\authorrunning{V. Sovrasov}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia \\
  \email{sovrasov.vlad@gmail.com}
}%
\maketitle              % typeset the header of the contribution
%

\begin{abstract}
In this paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University.
The Globalizer is designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. Also this work considers a method of
heuristic hyperparameters control for Globalizer allowing to reduce amount of
initial tuning before optimization. The proposed scheme allows rapidly increase convergence speed
of Globalizer by switching between ``local'' and ``global'' search phases in runtime.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

@@

Earlier, the comparison of the stochastic global optimization algorithms \cite{Ali2005, JSSv060i06}
as well as of the deterministic ones \cite{posik2012, KVASOV2018245, Liberti2005} between each
other has been considered in the literature. In these works, most of modern methods have
been studied in details, but the emphasis was made mainly on the sequential algorithms while the
reliability and the convergence speed were the main criteria of efficiency of the optimization
methods. In the majority of works, the sets of well-known test problems (for example, the Rastrigin
function, Ackley function, etc.) were taken as the sets of test functions. The sizes of such sets don't
exceed 100 different functions usually, some of which can be the single-extremal ones (such as the
Rosenbrock function).

@@

Optimization method applied in Globalizer to solve the reduced problem
(\ref{eq:oneDimTask}) is based on the AGS method, which can be presented as follows ---
see \cite{strongin1978}, \cite{strSergGO}.
\par
The algorithm considered for solving the stated problem implies generating
a sequence of points \(x_k\), in which the values of the minimized function \(z_k = f(x_k)\)
are computed. Let us call the process of computating the function value
(including calculating an image \(y^k=y(x^k)\)) a trial, and the pair \((x^k,z^k)\) ---
the result of the trial. A set of the pairs \(\{(x^k,z^k)\}, 1\leqslant k\leqslant n\)
makes up the search information accumulated by the method after executing \(n\) steps.
\par
The initial iteration of the algorithm is performed at an arbitrary point \mbox{\(x^1\in(0,1)\)}.
Then, let us suppose that \(k\), \(k\ge 1\), optimization iterations have been completed already.
The selection of the trial point \(x^{k+1}\) for the next iteration is performed according to the
following rules.

@@

\textit{Step 5.} Execute a new trial at point \(x_{k+1}\) computed according to the formula
\begin{displaymath}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2},t=1,t=k+1,
\end{displaymath}
\begin{displaymath}
\label{step5}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2}-\sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,1<t<k+1.
\end{displaymath}

@@

\subsubsection{Hyperparameters tuning}

The parameter $r$ from (\ref{eq:step2}) affects the global convergence of AGS directly (see
\cite{strSergGO}, Chapter 8):
at high enough value of $r$, the method converges to all global minima of the objective function with
guarantee.
At the same time, according to (\ref{eq:step3_2}), at the infinitely high value of $r$, AGS turns into
the brute force search method on a uniform grid.

@@

The use of the sets of test problems with known solutions generated by some random mechanisms is
one of commonly accepted approaches to comparing the optimization algorithms
\cite{Beiranvand2017}. In the present work, we will use two generators of test problems generating
the problems of different nature \cite{grishaginClass, Gaviano2003}.

@@

\section{Results of Numerical Experiments}

The results of various algorithms on different problem classes depend on the adjustments of
algorithms directly. In most cases, the authors of software implementations are oriented onto the
problems of medium difficulty. In order to obtain a satisfactory result when solving the essentially
multiextremal problems, a correction of some parameters is required. When conducting the
comparison, the following parameters for the methods were employed:
\begin{itemize}
  \item in the AGS-AR method, the parameter of alternation the
  global and local stages $q$ was set to be equal to $50\cdot\log_2(N-1)\cdot N^2$, also $r_{min}=3,\:r_{max}=2\cdot r_{min}$;
  \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
  \item in the SDA method, the parameter \(visit=2.72\).
\end{itemize}

@@

@@

In the present paper, several sequential and parallel global optimization algorithms were considered.
A comparison of efficiencies of these ones has been done on a set of test problems. The results allow
making the following conclusions:
\begin{itemize}
  \item the proposed modification of the stock AGS, AGS-AR allows to pay less attention to initial hyperparameter tuning and
  performs on-par with properly tuned AGS.
  \item AGS-AR method has demonstrated the convergence
  speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
  implementations of which are available;
  \item the stochastic optimization methods inferior to the deterministic ones in the convergence
speed and in reliability. It is manifested especially strongly on more complex multiextremal
problems;
\end{itemize}

>>>>

\begin{document}
%
\title{Comparison of Several Stochastic and Deterministic Derivative-free Global Optimization Algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Vladislav Sovrasov}
%
\authorrunning{V. Sovrasov}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia \\
  \email{sovrasov.vlad@gmail.com}
}%
\maketitle              % typeset the header of the contribution
%

\begin{abstract}
In this paper popular open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University.
The Globalizer is designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. Also this work considers a method of
heuristic hyperparameters control for the Globalizer allowing to reduce amount of
initial tuning before optimization. The proposed scheme allows substantially increase convergence speed
of the Globalizer by switching between ``local'' and ``global'' search phases in runtime.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ algorithms comparison $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization $\cdot$ multi-extremal problems}
\end{abstract}
%
\section{Introduction}

@@

Earlier, the comparison of the stochastic global optimization algorithms \cite{Ali2005, JSSv060i06}
as well as of the deterministic ones \cite{posik2012, KVASOV2018245, Liberti2005} between each
other has been considered in the literature. In these works, most of modern methods have
been studied in details. In the majority of works, the sets of well-known test problems (for example, the Rastrigin
function, Ackley function, etc.) were taken as the sets of test functions. The sizes of such sets don't
exceed 100 different functions usually, some of which can be the single-extremal ones (such as the
Rosenbrock function).

@@

Optimization method applied in Globalizer \cite{globalizerSystem} to solve the reduced problem
(\ref{eq:oneDimTask}) is based on the AGS method, which can be presented as follows ---
see \cite{strongin1978}, \cite{strSergGO}.
\par
The algorithm considered for solving the stated problem implies generating
a sequence of points \(x_k\), in which the values of the minimized function \(z_k = f(x_k)\)
are computed. Let us call the process of computating the function value
(including calculating an image \(y^k=y(x^k)\)) a trial, and the pair \((x^k,z^k)\) ---
the result of the trial. A set of the pairs \(\{(x^k,z^k)\}, 1\leqslant k\leqslant n\)
makes up the search information accumulated by the method after executing \(n\) steps.
\par
The initial iteration of the algorithm is performed at an arbitrary point \mbox{\(x^1\in(0,1)\)}.
Then, let us suppose that \(k\), \(k\ge 1\), optimization iterations have been completed already.
The selection of the trial point \(x^{k+1}\) for the next iteration is performed according to the
following rules.

@@

\textit{Step 5.} Execute a new trial at point \(x_{k+1}\) computed according to the formula
\begin{displaymath}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2},t=1,t=k+1,
\end{displaymath}
\begin{equation}
\label{step5}
x_{k+1}=\dfrac{x_{t}+x_{t-1}}{2}-\sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,1<t<k+1.
\end{equation}

@@

\subsubsection{Hyperparameters control in AGS}

The parameter $r$ from (\ref{eq:step2}) affects the global convergence of AGS directly (see
\cite{strSergGO}, Chapter 8):
at high enough value of $r$, the method converges to all global minima of the objective function with
guarantee.
At the same time, according to (\ref{eq:step3_2}) and (\ref{step5}), at the infinitely high value of $r$, AGS turns into
the brute force search method on a uniform grid.

@@

The use of the sets of test problems with known solutions generated by some random mechanisms is
one of commonly accepted approaches to comparing the optimization algorithms
\cite{Beiranvand2017}. In the present work, we will use two generators of test problems generating
the problems of different nature \cite{grishaginClass, Gaviano2003} \footnote{Software implementations of
these generators are available in source codes at the page \url{https://github.com/sovrasov/global-optimization-test-problems}}.

@@

The average number of trials doesn't represent the real behavior of an optimization method
on a problems set in some cases. For an instance, if a method performs well on the most problems
and spends too much trials to solve the least several problems, we wouldn't catch such
case looking at the average number of trials only.
As an advanced measure of performance we will use the operating characteristic \cite{grishaginClass}.
It's defined by a set of points on the \((K, P)\) plane where \(K\) is the average number of search trials
conducted before satisfying the termination condition when minimizing a function
from a given class, and \(P\) is the proportion of problems solved successfully.
If at a given \(K\), the operating characteristic of a method goes higher than one
from another method, it means that at fixed search costs, the former method has a
greater probability of finding the solution. If some value of \(P\) is fixed, and the
characteristic of a method goes to the left from that of another method, the former
method requires fewer resources to achieve the same reliability.

\section{Results of Numerical Experiments}
\label{sec:experiments}
The results of various algorithms on different problem classes depend on the adjustments of
algorithms directly. In most cases, the authors of software implementations are oriented onto the
problems of medium difficulty. In order to obtain a satisfactory result when solving the essentially
multiextremal problems, a correction of some parameters is required. When conducting the
comparison, the following parameters for the methods were employed:
\begin{itemize}
  \item in the AGS-AR method, the parameter of alternation the
  global and local stages $q$ was set to be equal to $50\cdot\log_2(N-1)\cdot N^2$, also $r_{min}=3,\:r_{max}=2\cdot r_{min}$;
  \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
  \item in the SDA method, the parameter \(visit=2.72\).
\end{itemize}

@@

\begin{figure}[ht]
  \centering
  \subfloat[4d Simple]{{\includegraphics[width=.5\textwidth]{images/gklss4d.pdf}}\label{fig:s4d}}
  \subfloat[4d Hard]{{\includegraphics[width=.5\textwidth]{images/gklsh4d.pdf}}\label{fig:h4d}}

  \subfloat[5d Simple]{{\includegraphics[width=.5\textwidth]{images/gklss5d.pdf}}\label{fig:s5d}}
  \subfloat[5d Hard]{{\includegraphics[width=.5\textwidth]{images/gklsh5d.pdf}}\label{fig:h5d}}
  \caption{Operating characteristics of the algorithms when solving problems from the GKLS 4d and 5d classes. Best viewed in color.}
\end{figure}

Operating characteristic of the methods (Figures \ref{fig:s4d}, \ref{fig:h4d}, \ref{fig:s5d}, \ref{fig:h5d})
demonstrates that AGS and AGS-AR faster than the other methods achieve 100\% success rate. Also on GKLS 5d Simple the DIRECT
generally has the best performance, but there are several hard problems that affect it's average number of
trials metric.

\paragraph{Robustness of AGS and AGS-AR to the Hyperparameters Choice.}

In order to investigate the influence of hyperparameters to the convergence speed of the AGS and AGS-AR,
experiments with the following settings were conducted on the problems from GKLS 5d Simple class:
\begin{itemize}
  \item AGS with $r=4$ (like in the Table \ref{tab:params});
  \item AGS with $r=6$;
  \item AGS-AR with parameters from the beginning of the Section \ref{sec:experiments}
  ($q=50\cdot\log_2(4)\cdot 25 = 2500$, $r_{min}=3,\:r_{max}=2\cdot r_{min}$);
  \item AGS-AR with $r_{max}=8$ and other parameters from the previous experiment;
  \item AGS-AR with $q=1000$ and other parameters from the beginning of the Section \ref{sec:experiments};
\end{itemize}

\begin{figure}[ht]
  \centering
  \includegraphics[width=.6\textwidth]{images/ar_stab.pdf}
  \caption{Operating characteristics of AGS and AGS-AR with different hyperparameters
  when solving problems from the GKLS 5d Simple classe. Best viewed in color.}
  \label{fig:stability}
\end{figure}

The operating characteristics collected in the experiments described above are shown in the Figure \ref{fig:stability}.
AGS with $r=6$ (the cyan-colored curve) shows the worst convergence speed, which indicates that AGS is very sensitive to choice of $r$.
Since on the start AGS-AR has the same value of $r$ as AGS with $r=6$, operating characteristics of these methods
are identical up to $K=2500$. After that point AGS-AR switches to $r=3$ and rapidly begins to increase the amount of solved problems
until the next exploration phase on $K=5000$. Intervals where AGS-AR works with $r=r_{max}$ are visible on the operating characteristics as plateaus.
Variations of $r$ and $q$ didn't drastically change the operating characteristic of AGS-AR. The latter observation shows robustness of the proposed
AGS modification with the alternating parameter $r$.

@@

In the present paper, several global optimization algorithms were considered.
A comparison of efficiencies of these ones has been done on a set of test problems.
Also a scheme of hyperparameters control for the AGS algorithms was proposed and evaluated.
The results presented in this work allow making the following conclusions:
\begin{itemize}
  \item the proposed modification of the stock AGS, AGS-AR allows to pay less attention to initial hyperparameter tuning and
  performs on-par with properly tuned AGS.
  \item AGS-AR method has demonstrated the convergence
  speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
  implementations of which are available;
  \item the stochastic optimization methods inferior to the deterministic ones in the convergence
speed and in reliability. It is manifested especially strongly on more complex multiextremal
problems;
\end{itemize}
